---
title: "Topic dashboard"
subtitle: "Script 4 of 4: build the standalone HTML dashboard"
format:
  typst:
    toc: false
execute:
  warning: false
  message: false
---

```{=typst}
#show figure: set block(breakable: true)
```

```{=html}
<!--
RUN ORDER (script 4 of 4; render any time after 02_topics.qmd)
  Rebuilds outputs/topic_dashboard.html from what scripts 2 and 3 wrote.
  It makes no API calls and fits nothing, so it takes seconds; rerun it
  after each 03_stance.qmd topic to add that topic's stance labels.
  The PDF this renders is only a build record. The HTML is the product.
-->
```

```{r config}
db <- list(
  title = "Topic dashboard",  # heading shown at the top of the page
  stance_rater = "stance_1"  # which rater's labels the dashboard shows
)

suppressPackageStartupMessages({
  library(here)
  library(tidyverse)  # tidyverse LAST so its verbs win
})
# Used namespaced: jsonlite, glue.
options(knitr.table.max.print = Inf, knitr.kable.max_rows = Inf)
`%||%` <- function(x, y) if (is.null(x)) y else x
p_out <- function(f) here::here("outputs", f)
source(here::here("code", "llm_source.R"))
```

# Inputs

The dashboard is built only from files already on disk. Each is required except the stance files, which appear one per topic as `03_stance.qmd` is run. The codebook is checked against the topic proportions it was drafted from: if `02_topics.qmd` has refit since, topics may be renumbered, and a stale codebook would put its labels on the wrong topics. The render stops rather than publish that.

```{r inputs}
#| results: asis
needed <- p_out(c("paragraph_theta.csv", "topic_codebook.csv",
                  "topic_summary.csv", "topic_trends.csv"))
invisible(fingerprint(needed))  # stops by name if any input is missing
check_fingerprint(p_out("topic_codebook.csv"), p_out("paragraph_theta.csv"),
                  strict = FALSE)

paragraph_theta <- readr::read_csv(needed[1], show_col_types = FALSE)
codebook <- readr::read_csv(needed[2], show_col_types = FALSE) |>
  dplyr::left_join(
    readr::read_csv(needed[3], show_col_types = FALSE) |>
      dplyr::select(topic, frex),
    by = "topic")
trends <- readr::read_csv(needed[4], show_col_types = FALSE)

stance_files <- list.files(here::here("outputs"),
                           pattern = "^stance_labels_topic\\d+\\.csv$",
                           full.names = TRUE)
stance_all <- stance_files |>
  purrr::map(readr::read_csv, show_col_types = FALSE) |>
  purrr::list_rbind()
stance <- if (nrow(stance_all) == 0) NULL else {
  s <- stance_all |> dplyr::filter(rater == db$stance_rater, !is.na(label))
  purrr::set_names(s$label, s$para_uid)
}

writeLines(glue::glue("{nrow(paragraph_theta)} paragraphs, {nrow(codebook)} \\
  topics, trends for {dplyr::n_distinct(trends$topic)} topics, stance labels \\
  from {length(stance_files)} topic file(s)."))
```

# Build

```{r build}
#| results: asis
out <- p_out("topic_dashboard.html")
write_dashboard(
  path = out,
  template = here::here("code", "dashboard_template.html"),
  paragraph_theta = paragraph_theta,
  codebook = codebook,
  trends = trends,
  stance = stance,
  title = db$title)

tibble::tibble(
  item = c("file", "size (MB)", "topics", "paragraphs", "documents",
           "topics with a proposition", "paragraphs with a stance label",
           "built"),
  value = c(basename(out), round(file.size(out) / 1e6, 2), nrow(codebook),
            nrow(paragraph_theta), dplyr::n_distinct(paragraph_theta$atom_id),
            sum(!is.na(codebook$proposition)), length(stance),
            format(Sys.time(), "%Y-%m-%d %H:%M"))) |>
  tbl_paged(caption = "Dashboard build record.")
```

# Before sharing it

The dashboard is one self-contained file: it opens in any browser, from disk or a web host, with no server and no internet connection. **It embeds the full text of every document**, so publishing it republishes the corpus; confirm that is acceptable for this corpus before posting it anywhere public. The prevalence chart is a model estimate with an interval and the yearly counts are exact, and the page labels them as the different claims they are.
