---
title: "Stance on a claim over time (experimental)"
subtitle: "Step 3 of 3. Reads preprocess.qmd outputs; does not need topics"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r config}
st <- list(
  name = "world_role",  # short name for this claim's output files
  proposition = "The United States should take an active role in the affairs of other nations.",
  negation = "The United States should stay out of the affairs of other nations and avoid foreign commitments.",
  strata = c(high = 0.05, mid = 0.20),  # top 5% by similarity, then the next 20%; the rest is low
  n = c(high = 200L, mid = 70L, low = 30L),  # calls per stratum; unused calls roll down
  seed = 2026L,
  window_n = 300L,  # calls per window
  window_wait = 60L,  # seconds between windows; 4 * 3600 for a rate-limited large model
  df_spline = 2L,  # flexibility of the trend
  min_class_n = 5L,  # paragraphs needed on each side for any trend
  min_year = NULL,  # optional: fit trends from this year on
  use_openrouter = TRUE,  # FALSE at work
  steps = tibble::tribble(
    ~step, ~base_url_env, ~api_key_env, ~model,
    "embed", "COMPASS_EMBED_URL", "COMPASS_EMBED_KEY", "MODEL-FOR-EMBEDDING",
    "stance", "COMPASS_SMALL_URL", "COMPASS_SMALL_KEY", "SMALL-MODEL"))

suppressPackageStartupMessages({
  library(here); library(splines); library(viridis)
  library(tidyverse)  # tidyverse last so its verbs win
})
theme_set(theme_minimal(base_size = 12))
source(here::here("code", "llm_source.R"))
if (isTRUE(st$use_openrouter)) {
  source(here::here("code", "llm_openrouter.R"))
  st$steps$model <- open_router_models[st$steps$step]
}
```

# The design

Coding every paragraph would cost one model call each. Instead, every paragraph is scored for free by how close its meaning is to the claim, and a sample is coded: heavier where the claim is likely to appear, never zero anywhere. This is two-phase stratified sampling. Similarity is the first-phase measure and the model's label the second, and each coded paragraph carries the weight N/n of its stratum, so estimates describe the whole corpus rather than only the paragraphs that were read.

1. **Similarity.** The claim and its negation are embedded with the same model as the corpus (2 embedding calls). Each paragraph gets the higher of its two cosines, so paragraphs that argue against the claim in their own words rank as high as ones that echo it.
2. **Strata.** Paragraphs are ranked by similarity and split into high, middle, and low strata.
3. **Coding.** A sample from each stratum is labeled favor, neutral, oppose, or irrelevant, one call per paragraph. The low stratum doubles as the audit of what the similarity screen misses.

**Claim:** `r st$proposition`

**Negation used for scoring:** `r st$negation`

```{r similarity}
#| results: asis
para_in <- p_out(c("paragraphs.csv", "paragraph_vectors.rds"))
paras <- readr::read_csv(para_in[1], show_col_types = FALSE)
V <- readRDS(para_in[2])[paras$para_uid, , drop = FALSE]
q <- embed_texts(c(st$proposition, st$negation), st$steps, st$use_openrouter)
if (ncol(q) != ncol(V))
  stop("The claim was embedded with a different model than the corpus. ",
       "Use the same embed row as preprocess.qmd.")
unit <- function(m) m / sqrt(rowSums(m^2))
sims <- unit(V) %*% t(unit(q))
paras <- paras |>
  dplyr::mutate(sim = pmax(sims[, 1], sims[, 2]),
                pct = dplyr::percent_rank(dplyr::desc(sim)),
                stratum = dplyr::case_when(
                  pct < st$strata[["high"]] ~ "high",
                  pct < st$strata[["high"]] + st$strata[["mid"]] ~ "mid",
                  TRUE ~ "low"))
writeLines(glue::glue("{nrow(paras)} paragraphs from \\
  {dplyr::n_distinct(paras$atom_id)} comments scored against the claim."))
```

```{r sample}
#| results: asis
# A stratum smaller than its allocation is coded in full and its unused
# calls roll down to the next stratum (high, then mid, then low), so the
# whole budget is spent.
N <- table(factor(paras$stratum, levels = c("high", "mid", "low")))
alloc <- purrr::reduce(c("high", "mid", "low"), function(acc, h) {
  want <- st$n[[h]] + acc$spare
  take <- min(N[[h]], want)
  list(n = c(acc$n, stats::setNames(take, h)), spare = want - take)
}, .init = list(n = integer(), spare = 0))$n
sampled <- paras |>
  dplyr::group_split(stratum) |>
  purrr::map(function(d) {
    n_h <- alloc[[d$stratum[1]]]
    i <- withr::with_seed(st$seed, sample.int(nrow(d), n_h))
    dplyr::mutate(d[sort(i), ], N_h = nrow(d), n_h = n_h, w = N_h / n_h)
  }) |>
  purrr::list_rbind()
sampled |>
  dplyr::group_by(stratum) |>
  dplyr::summarise(paragraphs = dplyr::first(N_h), coded = dplyr::first(n_h),
                   weight = round(dplyr::first(w), 1)) |>
  dplyr::left_join(paras |>
                     dplyr::group_by(stratum) |>
                     dplyr::summarise(similarity = str_c(round(min(sim), 2), " to ",
                                                         round(max(sim), 2))),
                   by = "stratum") |>
  dplyr::arrange(match(stratum, c("high", "mid", "low"))) |>
  dplyr::select(stratum, similarity, paragraphs, coded, weight) |>
  tbl_paged(caption = "Strata. Each coded paragraph stands for `weight` paragraphs of its stratum.")
```

# Coding

One paragraph per call, at temperature 0, without the year. Labels are saved per paragraph after every window of `r st$window_n` calls, so an interrupted run resumes where it stopped, and a changed sample only sends paragraphs not yet coded. Changing the claim's wording refuses the saved labels and names the file to delete.

```{r classify}
#| results: asis
labels_path <- p_out(str_c("stance_labels_", st$name, ".csv"))
prompt_path <- p_out(str_c("stance_prompt_", st$name, ".txt"))
sys <- build_prompt(
  role = "You are a survey methodologist coding stance in public comments.",
  task = str_c("Decide the author's position toward this proposition: \"", st$proposition,
               "\". First decide whether the text addresses the proposition at all; ",
               "if not, the label is irrelevant. Otherwise decide favor, oppose, or neutral."),
  rules = c("Stance is measured only toward the proposition.",
            "Stance is not tone: an angry passage can favor the proposition.",
            "Judge only the text shown; use no outside knowledge.",
            "If the text is ambiguous, choose neutral rather than guessing."),
  output = "One label: favor, neutral, oppose, or irrelevant.")
writeLines(sys, prompt_path)
check_inputs(labels_path, c(para_in[1], prompt_path))

done <- if (file.exists(labels_path))
  readr::read_csv(labels_path, show_col_types = FALSE, col_types = "cc") else
  tibble(para_uid = character(), label = character())
todo <- dplyr::anti_join(sampled, done, by = "para_uid")
windows <- split(todo, ceiling(seq_len(nrow(todo)) / st$window_n))
type <- ellmer::type_object(label = ellmer::type_enum(
  c("favor", "neutral", "oppose", "irrelevant"), "The single best label."))

done <- purrr::reduce(seq_along(windows), function(done, i) {
  if (i > 1) purrr::walk(seq_len(st$window_wait), function(s) Sys.sleep(1),
                         .progress = glue::glue("resting {st$window_wait}s"))
  d <- windows[[i]]
  message(glue::glue("window {i} of {length(windows)}: {nrow(d)} calls"))
  res <- ask_many(make_chat("stance", sys, st$steps, st$use_openrouter),
                  str_c("Text: ", d$text), type, rpm = st$window_n)
  if (all(is.na(res$label)))
    stop("Every call in the window failed. First error: ",
         conditionMessage(purrr::compact(res$.error)[[1]]))
  out <- dplyr::bind_rows(done, tibble(para_uid = d$para_uid,
                                       label = as.character(res$label))) |>
    dplyr::filter(!is.na(label))
  readr::write_csv(out, labels_path)
  record_inputs(labels_path, c(para_in[1], prompt_path))
  out
}, .init = done)

coded <- dplyr::inner_join(sampled, done, by = "para_uid")
if (nrow(coded) < nrow(sampled))
  writeLines(glue::glue("**{nrow(sampled) - nrow(coded)} calls failed;** re-render to retry them."))
coded |>
  dplyr::count(stratum, label) |>
  tidyr::pivot_wider(names_from = label, values_from = n, values_fill = 0) |>
  dplyr::left_join(coded |>
                     dplyr::group_by(stratum) |>
                     dplyr::summarise(relevant_share = round(mean(label != "irrelevant"), 2),
                                      weight = round(dplyr::first(w), 1)),
                   by = "stratum") |>
  dplyr::arrange(match(stratum, c("high", "mid", "low"))) |>
  tbl_paged(caption = "Labels by stratum (counts). relevant_share in the low stratum is the audit: how often the similarity screen ranked a relevant paragraph low. Each of those stands for `weight` paragraphs, so a high value there widens every interval.")
```

# What the labels look like

```{r examples}
#| results: asis
coded |>
  dplyr::group_by(label) |>
  dplyr::slice_max(sim, n = 3, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(title = str_c(toupper(label), " | ", Year, " | ", para_uid),
                text = str_trunc(text, 350)) |>
  tbl_records("title", "text",
              caption = "The three paragraphs most similar to the claim for each label. If these do not match what the label should mean, fix the claim before reading any number.")
```

# Results

All shares are weighted to the whole corpus, with 95 percent intervals clustered by comment.

```{r overall}
#| results: asis
rel <- dplyr::filter(coded, label != "irrelevant")
sided <- dplyr::filter(rel, label %in% c("favor", "oppose"))
dplyr::bind_rows(
  dplyr::mutate(share_ci(coded$label != "irrelevant", coded$atom_id, coded$w),
                measure = "paragraphs addressing the claim", coded = nrow(coded)),
  dplyr::mutate(share_ci(rel$label == "favor", rel$atom_id, rel$w),
                measure = "favor, of those addressing it", coded = nrow(rel)),
  dplyr::mutate(share_ci(rel$label == "neutral", rel$atom_id, rel$w),
                measure = "neutral, of those addressing it", coded = nrow(rel)),
  dplyr::mutate(share_ci(sided$label == "favor", sided$atom_id, sided$w),
                measure = "favor, of those taking a side", coded = nrow(sided))) |>
  dplyr::select(measure, coded, share, lo, hi) |>
  tbl_paged(caption = "Weighted shares for the whole corpus.")
```

Two lines over time. **Stance** is the share of paragraphs addressing the claim that favor it. **Attention** is the share of all paragraphs that address the claim at all. Read together, they separate a change in opinion from a change in how often the claim comes up.

```{r trends}
#| results: asis
from <- st$min_year %||% min(coded$Year)
fit_one <- function(d, y, what) {
  d <- dplyr::filter(d, Year >= from)
  r <- tryCatch(trend_curve(y(d), d$Year, d$atom_id, st$df_spline,
                            st$min_class_n, d$w),
                error = function(e) conditionMessage(e))
  if (is.character(r)) {
    writeLines(glue::glue("**{what}: no trend fitted:** {r}."))
    return(NULL)
  }
  if (r$df[1] == 1L) writeLines(glue::glue("{what}: fewer than 10 paragraphs on one side, so a straight line in year."))
  dplyr::mutate(r, line = what)
}
rug <- dplyr::bind_rows(
  dplyr::transmute(dplyr::filter(rel, Year >= from), Year,
                   line = "stance: favor, of those addressing the claim"),
  dplyr::transmute(dplyr::filter(coded, Year >= from), Year,
                   line = "attention: share addressing the claim"))
trends <- dplyr::bind_rows(
  fit_one(rel, function(d) as.integer(d$label == "favor"),
          "stance: favor, of those addressing the claim"),
  fit_one(coded, function(d) as.integer(d$label != "irrelevant"),
          "attention: share addressing the claim"))
if (nrow(trends)) readr::write_csv(trends, p_out(str_c("stance_trend_", st$name, ".csv")))
```

```{r trend-plot}
#| fig-height: 6
#| fig-cap: "Weighted trends with 95 percent intervals clustered by comment. Ticks along the bottom mark the coded paragraphs behind each line; where they are sparse, the line is mostly the model's smoothing. The intervals do not include error in the model's labels."
if (nrow(trends))
  ggplot(trends, aes(Year, fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = viridis(1, begin = 0.35)) +
    geom_line(linewidth = 0.9, color = viridis(1, begin = 0.35)) +
    geom_rug(data = dplyr::semi_join(rug, trends, by = "line"), aes(Year),
             inherit.aes = FALSE, alpha = 0.4, length = grid::unit(0.03, "npc")) +
    facet_wrap(~ line, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = function(x) str_c(round(100 * x), "%")) +
    labs(x = NULL, y = NULL)
```

# Before using these numbers

This step is experimental. Labels come from one model with no human check, so their error is not in the intervals. Relevant paragraphs found in the low stratum carry large weights and widen the intervals; if many turn up there, move budget from the high stratum to the middle. Every share is a share of paragraphs, so a long comment counts more than a short one.
