---
title: "Stance on a claim over time"
subtitle: "Step 3 of 3. Reads topics.qmd outputs"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

```{r config}
st <- list(
  name = "world_role",  # short name for this claim's output files
  proposition = "The United States should take an active role in the affairs of other nations.",
  df_spline = 2L,  # flexibility of the trends
  min_class_n = 5L,  # paragraphs a label needs, in and out, for a trend
  save_every = 100L)  # labels are saved after this many paragraphs

suppressPackageStartupMessages({
  library(here); library(splines)
  library(tidyverse)  # tidyverse last so its verbs win
})
source(here::here("code", "llm_source.R"))  # also sources app/R/stance_core.R
ep <- stance_endpoint()  # STANCE_URL and STANCE_MODEL from .Renviron
```

# The design

Every paragraph in the corpus is read by a language model and labeled for its stance toward one claim. Nothing is sampled, so there is no sampling error and no weights: the uncertainty left is the year-to-year variation of the trend and the error in the labels themselves.

1. **Coding.** Each paragraph goes to the model alone, one request at a time, at temperature 0, without its year or its comment. The model first decides whether the paragraph addresses the claim at all (otherwise *irrelevant*), then whether it favors the claim, opposes it, or is neutral.
2. **Trends.** For each label, its share of all paragraphs written in a year is modeled as a smooth function of year (a logistic regression on a natural spline). Paragraphs from one comment are related, so the 95 percent intervals are clustered by comment.

The denominator is every coded paragraph. That keeps every year's full count behind each interval, but it means a rising *favor* line can come from more writers favoring the claim or from more writers discussing it at all. The *irrelevant* line shows which: if it falls while *favor* rises, the claim is being discussed more.

The app's Stance tab runs these same functions, so a claim submitted there gets the same treatment as the one in this report.

**Claim:** `r st$proposition`

**Model:** `r ep$model`

# Coding

Labels are saved every `r st$save_every` paragraphs, so an interrupted run resumes where it stopped. Changing the claim or the model refuses the saved labels and names the file to delete.

```{r classify}
#| results: asis
theta_in <- p_out("paragraph_theta.csv")
paras <- readr::read_csv(theta_in, show_col_types = FALSE) |>
  dplyr::select(para_uid, atom_id, Year, text)
labels_path <- p_out(str_c("stance_labels_", st$name, ".csv"))
prompt_path <- p_out(str_c("stance_prompt_", st$name, ".txt"))
writeLines(c(stance_prompt(st$proposition), "", str_c("MODEL ", ep$model)), prompt_path)
check_inputs(labels_path, c(theta_in, prompt_path))

done <- if (file.exists(labels_path))
  readr::read_csv(labels_path, show_col_types = FALSE, col_types = "cc") else
  tibble(para_uid = character(), label = character())
todo <- dplyr::anti_join(paras, done, by = "para_uid")
pb <- utils::txtProgressBar(max = max(1, nrow(todo)), style = 3, file = stderr())
chunks <- split(todo, ceiling(seq_len(nrow(todo)) / st$save_every))
done <- purrr::reduce(seq_along(chunks), function(done, i) {
  d <- chunks[[i]]
  offset <- (i - 1) * st$save_every
  lab <- code_all(d$text, st$proposition, ep, on_step = function(j, n)
    utils::setTxtProgressBar(pb, offset + j))
  out <- dplyr::bind_rows(done, tibble(para_uid = d$para_uid, label = lab)) |>
    dplyr::filter(!is.na(label))
  readr::write_csv(out, labels_path)
  record_inputs(labels_path, c(theta_in, prompt_path))
  out
}, .init = done)
close(pb)

coded <- dplyr::inner_join(paras, done, by = "para_uid")
if (nrow(coded) < nrow(paras))
  writeLines(glue::glue("**{nrow(paras) - nrow(coded)} paragraphs got no label;** re-render to retry them."))
dplyr::count(coded, label) |>
  dplyr::mutate(share = n / sum(n)) |>
  tbl_paged(caption = glue::glue("Labels for {nrow(coded)} paragraphs from {dplyr::n_distinct(coded$atom_id)} comments."))
```

# What the labels look like

```{r examples}
#| results: asis
withr::with_seed(2026L, coded |>
  dplyr::filter(label != "irrelevant") |>
  dplyr::group_by(label) |>
  dplyr::slice_sample(n = 3) |>
  dplyr::ungroup()) |>
  dplyr::mutate(title = str_c(toupper(label), " | ", Year, " | ", para_uid),
                text = str_trunc(text, 350)) |>
  tbl_records("title", "text",
              caption = "Three paragraphs drawn at random for each label. If these do not match what the label should mean, fix the claim before reading any number.")
```

# Results

```{r overall}
#| results: asis
stance_overall(coded$label, coded$atom_id) |>
  tbl_paged(caption = "Share of all paragraphs with each label, with 95 percent intervals clustered by comment.")
```

```{r trends}
#| results: asis
trends <- stance_trends(coded$label, coded$Year, coded$atom_id, st$df_spline, st$min_class_n)
purrr::walk(attr(trends, "notes"), function(x) writeLines(str_c("**", x, "**\n")))
if (any(trends$df == 1L))
  writeLines(glue::glue("Straight line in year for: {str_c(unique(trends$label[trends$df == 1L]), collapse = ', ')} (fewer than 10 paragraphs with the label)."))
stance_record(trends, st$proposition, ep, nrow(coded)) |>
  readr::write_csv(p_out(str_c("stance_trend_", st$name, ".csv")))
```

```{r trend-plot}
#| fig-cap: "Each label's share of all paragraphs by year, with 95 percent intervals clustered by comment. Each panel has its own scale. The intervals do not include error in the model's labels."
stance_plot(trends)
```

# Before using these numbers

Labels come from one model with no human check, so their error is not in the intervals; read the examples above and re-run with a paraphrase of the claim to see whether the result turns on its wording. Every share is a share of paragraphs, so a long comment counts more than a short one.
