---
title: "Public comments: paragraph corpus and structural topic model"
subtitle: "Script 1 of 2: text mining, STM, modal topic assignment"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

```{=typst}
#show figure: set block(breakable: true)
```

```{=html}
<!--
PRODUCTION CHECKLIST (simulation -> real data)
  1. Replace the Corpus chunk with your own read:
       corpus_raw <- readr::read_csv(here("input", "YOUR_FILE.csv"))
     Create text_mining/input/ and drop the CSV there; the file name is the
     only blank. Required columns: atom_id (document id), Year (integer),
     body_english. Any source is fine: the validation gate runs on whatever
     filled corpus_raw. Retune tx$K_grid for the larger corpus.
  2. Read the newline diagnostic; set tx$para_split if paragraphs are
     blank-line separated. Delete outputs/ ONCE (caches track files, not code).
  3. Render with tx$run_searchK = TRUE (hours), read the diagnostics, set
     tx$K, set tx$run_searchK = FALSE, re-render.
  4. Script 2 reads outputs/. Nothing else here changes between projects.
OUTPUTS: tabular -> .csv, model objects and lists -> .rds, all in outputs/.
-->
```

```{r config}
# The only block you edit. Every analyst decision is a line here.
tx <- list(
  seed = 2026L,
  para_split = "\\r?\\n+",  # blank-line separated? "\\r?\\n{2,}"
  para_min_tokens = 15L,  # body fragments below this merge BACKWARD
  heading_max_tokens = 12L,  # headings merge FORWARD (see atomization)
  min_docfreq = 5L,  # DTM vocabulary floor
  run_searchK = TRUE,  # TRUE for the long K-selection run
  K_grid = seq(5, 30, 3),  # widen for a larger corpus
  K = 12L,  # set from the searchK diagnostics
  spline_df = 5L,
  facets_per_page = 9L,  # topic trend panels per page
  workers = 16L,

  # --- LLM topic labeling (runs once, after the final STM) -------------------
  use_openrouter = TRUE,  # FALSE on the work server
  # One row per step. base_url_env and api_key_env are the NAMES of the
  # .Renviron variables holding that step's endpoint and key, so the large
  # model and the stance models can use different credentials and nothing
  # sensitive appears in this file. Ignored when use_openrouter = TRUE.
  steps = tibble::tribble(
    ~step,     ~base_url_env,       ~api_key_env,        ~model,
    "label",   "COMPASS_SMALL_URL", "COMPASS_SMALL_KEY", "MODEL-FOR-LABELING",
    "manager", "COMPASS_LARGE_URL", "COMPASS_LARGE_KEY", "MODEL-FOR-MANAGER")
)

suppressPackageStartupMessages({
  library(here); library(stm); library(quanteda); library(tidytext)
  library(splines)
  library(furrr); library(viridis)
  library(tidyverse)   # tidyverse LAST so its verbs win
})
theme_set(theme_minimal(base_size = 12))
# Tables must not silently lose rows in the rendered PDF: knitr truncates long
# tables by default, and a 25-topic summary is past that limit.
options(knitr.table.max.print = Inf, knitr.kable.max_rows = Inf,
        max.print = 1e6)
set.seed(tx$seed)
# All paths are anchored with here(), so chunks run the same whether you
# render from the project root or run them interactively from code/.
# Layout: here() is the text_mining folder, holding code/, input/, outputs/.
`%||%` <- function(x, y) if (is.null(x)) y else x
p_out <- function(f) here::here("outputs", f)
source(here::here("code", "llm_source.R"))
if (isTRUE(tx$use_openrouter)) {
  source(here::here("code", "llm_use_openrouter_openrouter.R"))
  tx$steps$model <- use_openrouter_models[tx$steps$step]
}
dir.create(here::here("outputs"), showWarnings = FALSE, recursive = TRUE)

# Checkpoints: compute once, reload after. Delete the file to re-fit.
cache_rds  <- function(path, expr) {
  if (file.exists(path)) return(readRDS(path))
  x <- force(expr); saveRDS(x, path); x
}
cache_csv <- function(path, expr) {
  if (file.exists(path)) return(readr::read_csv(path, show_col_types = FALSE))
  x <- force(expr); readr::write_csv(x, path); x
}
```

# Corpus

The default corpus is `quanteda::data_corpus_inaugural`: 60 US inaugural addresses, 1789 to 2025, with `Year` already carried as a docvar. It ships with quanteda, so there is no extra dependency, and it is a real corpus of multi-paragraph documents rather than a simulation, which means the atomizer, the topic model, and the LLM codebook all run against text that behaves like text.

Two properties worth knowing before you read anything into the results. It yields about 1,874 paragraphs, roughly a tenth of a production corpus, so `tx$K` belongs in the low teens rather than the mid twenties. And it has **one document per year**, so document clusters and year strata coincide; that is fine here, but Script 2's cluster-robust variance cannot be exercised on it, since your production corpus has many comments per year and this one has exactly one.

**AT WORK** replace this chunk with the read of your own table. Everything downstream is unchanged; the validation gate below runs on whatever filled `corpus_raw`.

```{r corpus}
corpus_raw <- tibble(
  atom_id = quanteda::docnames(quanteda::data_corpus_inaugural),
  Year = quanteda::docvars(quanteda::data_corpus_inaugural, "Year"),
  body_english = as.character(quanteda::data_corpus_inaugural))

# AT WORK, this chunk becomes one line:
# corpus_raw <- readr::read_csv(here("input", "YOUR_FILE.csv"))
# Required columns: atom_id (document id), Year (integer), body_english (text).
```

# Corpus validation (keep this at work)

Stops the render with a named error on a bad read, which is cheaper than debugging a silent failure five chunks later.

```{r validate}
miss <- setdiff(c("atom_id", "Year", "body_english"), names(corpus_raw))
if (length(miss)) stop("corpus_raw missing column(s): ",
                       str_c(miss, collapse = ", "))
stopifnot(
  "atom_id must be unique"     = !anyDuplicated(corpus_raw$atom_id),
  "Year must be integer-valued" = all(corpus_raw$Year ==
                                        as.integer(corpus_raw$Year)),
  "body_english must be character" = is.character(corpus_raw$body_english))
n0 <- nrow(corpus_raw)
corpus_raw <- corpus_raw |>
  dplyr::filter(!is.na(Year), !is.na(body_english),
                str_squish(body_english) != "")
glue::glue("Validation: kept {nrow(corpus_raw)} of {n0} comments, ",
           "{min(corpus_raw$Year)}-{max(corpus_raw$Year)}")
```

# Newline diagnostic

The atomizer can only split on what the source uses to mark paragraphs, and Excel-origin text (in-cell breaks survive as `\n`) carries three conventions needing three responses. Rules: mostly line breaks present, keep the default; blank lines dominant **and** the visible sample shows single breaks used as soft wraps, set `tx$para_split = "\\r?\\n{2,}"`; almost no newlines, stop, since sentence chunking changes the measurement unit and is a design decision.

```{r newline-diagnostic}
#| results: asis
corpus_raw |>
  dplyr::mutate(b = str_replace_all(body_english, "\r\n", "\n")) |>
  dplyr::summarise(comments = dplyr::n(),
                   zero_newlines   = sum(str_count(b, "\n") == 0),
                   has_blank_lines = sum(str_count(b, "\n{2,}") > 0),
                   median_breaks   = median(str_count(b, "\n"))) |>
  tbl_paged(caption = "Line-break conventions; apply the decision rules above before the long render.")

corpus_raw |> dplyr::slice_sample(n = 2) |> dplyr::pull(body_english) |>
  str_replace_all("\r?\n", "<NL>\n") |>
  str_c(collapse = "\n\n==== NEXT ====\n\n") |> writeLines()
```

# Paragraph atomization

The unit of analysis is the paragraph: a comment spans several topics, and a topic model's "document" is whatever unit the DTM rows are. Merging is **direction-aware**, because short blocks come in two kinds that belong in opposite directions. Headings (at most `r tx$heading_max_tokens` tokens, no sentence-ending punctuation) merge **forward** into the paragraph they introduce, since a naive rule would glue a subtitle's words onto the *preceding* paragraph and misfile its topic signal; a trailing heading falls back to the last body paragraph. Short body fragments that do end like sentences merge **backward** into what they continue, absorbing codas, salutations, and signature lines.

```{r atomize}
count_tokens <- function(x) str_count(x, "\\S+")

paragraphs <- cache_csv(p_out("paragraphs.csv"), {
  # PASS 1 (forward): rev(cumsum(rev(!heading))) gives each body block an id
  # that decreases down the document and gives a heading the id of the NEXT
  # body block; trailing headings get 0, folded in by pmax(1L).
  corpus_raw |>
    dplyr::mutate(para_raw = str_split(body_english, tx$para_split)) |>
    dplyr::select(atom_id, Year, para_raw) |>
    tidyr::unnest(para_raw) |>
    dplyr::mutate(para_raw = str_squish(para_raw)) |>
    dplyr::filter(para_raw != "") |>
    dplyr::group_by(atom_id, Year) |>
    dplyr::mutate(
      heading = count_tokens(para_raw) <= tx$heading_max_tokens &
                !str_detect(para_raw, "[.!?]['\")\\]]*$"),
      fwd     = rev(cumsum(rev(!heading))) |> pmax(1L)) |>
    dplyr::group_by(atom_id, Year, fwd) |>
    dplyr::summarise(para_raw = str_c(para_raw, collapse = " "),
                     .groups = "drop") |>
    dplyr::arrange(atom_id, Year, dplyr::desc(fwd)) |>
    # PASS 2 (backward): short body fragments join the preceding paragraph.
    dplyr::group_by(atom_id, Year) |>
    dplyr::mutate(full = count_tokens(para_raw) >= tx$para_min_tokens,
                  blk  = cummax(cumsum(full) * full) |> pmax(1L)) |>
    dplyr::group_by(atom_id, Year, blk) |>
    dplyr::summarise(text = str_c(para_raw, collapse = " "),
                     .groups = "drop_last") |>
    dplyr::mutate(para_id  = dplyr::row_number(),
                  para_uid = str_c(atom_id, "-", sprintf("%02d", para_id)),
                  n_tokens = count_tokens(text)) |>
    dplyr::ungroup() |>
    dplyr::filter(n_tokens >= tx$para_min_tokens) |>
    dplyr::select(para_uid, atom_id, para_id, Year, n_tokens, text)
})
glue::glue("{nrow(paragraphs)} paragraphs from ",
           "{dplyr::n_distinct(paragraphs$atom_id)} comments")
```

# Corpus description

```{r eda}
#| fig-cap: "Paragraph volume by year: the denominator for every time-indexed estimate. Thin years mean the prevalence spline is interpolating and single-year claims are weak."
paragraphs |>
  dplyr::count(Year) |>
  ggplot(aes(Year, n)) +
  geom_col(fill = viridis(1, begin = 0.35)) +
  labs(x = NULL, y = "paragraphs")

tokens <- paragraphs |>
  tidytext::unnest_tokens(word, text) |>
  dplyr::anti_join(tidytext::get_stopwords(), by = "word") |>
  dplyr::filter(str_detect(word, "[a-z]"))
```

# Document-term matrix

Rare terms (under `r tx$min_docfreq` paragraphs) are trimmed to stabilize estimation. `meta` is built from the surviving dfm rownames, so alignment with the model survives any dropped document. At work, check `nfeat(dfm_para)`: a vocabulary in the low thousands is comfortable at this scale, and a much larger one argues for a higher floor.

```{r dtm}
dfm_para <- tokens |>
  dplyr::count(para_uid, word) |>
  tidytext::cast_dfm(para_uid, word, n) |>
  quanteda::dfm_trim(min_docfreq = tx$min_docfreq)
stm_in <- quanteda::convert(dfm_para, to = "stm")
# as.data.frame, not a tibble: stm subsets metadata with base [ , ] semantics
# when building the prevalence model matrix, and a tibble returns a column
# rather than a vector there ("Error creating model matrix").
meta   <- tibble(para_uid = rownames(dfm_para)) |>
  dplyr::left_join(paragraphs, by = "para_uid") |>
  as.data.frame()
stopifnot("every dfm row must match a paragraph" = !anyNA(meta$Year),
          "meta must align with the stm documents" =
            nrow(meta) == length(stm_in$documents))

# Spline basis as ORDINARY COLUMNS, so no function call ever appears inside a
# formula. stm builds its prevalence model matrix by evaluating the formula
# against the metadata, and a call there (s(), ns(), or anything namespaced)
# is the documented cause of "Error creating model matrix". prev_form then
# reads ~ yr1 + yr2 + ..., which is also what estimateEffect uses below.
bs <- splines::ns(meta$Year, df = tx$spline_df)
colnames(bs) <- str_c("yr", seq_len(ncol(bs)))
meta <- cbind(meta, as.data.frame(bs))
prev_rhs  <- str_c(colnames(bs), collapse = " + ")   # "yr1 + yr2 + ..."
prev_form <- as.formula(str_c("~ ", prev_rhs))
```

# Choosing K

stm's own parallelism forks, which is unavailable on Windows, so the held-out grid is parallelized explicitly with furrr. Read the panels together: take the smallest K where held-out likelihood has flattened and residual dispersion has stopped falling, then break ties on the coherence-exclusivity frontier. If the frontier is still improving at the top of the grid, widen `tx$K_grid` rather than extrapolating. Semantic coherence is $\sum_{i<j}\log\{(D(v_i,v_j)+1)/D(v_i)\}$ over each topic's top words, with the denominator the higher-ranked word.

```{r searchK}
if (isTRUE(tx$run_searchK)) {
  future::plan(future::multisession, workers = tx$workers)
  ho <- stm::make.heldout(stm_in$documents, stm_in$vocab, seed = tx$seed)
  metrics <- cache_csv(p_out("searchK_metrics.csv"), {
    furrr::future_map(tx$K_grid, function(k) {
      fit <- stm::stm(ho$documents, ho$vocab, K = k, init.type = "Spectral",
                      prevalence = prev_form, data = meta, verbose = FALSE)
      tibble(K = k,
             heldout   = stm::eval.heldout(fit, ho$missing)$expected.heldout,
             dispersion = stm::checkResiduals(fit, ho$documents)$dispersion,
             coherence = mean(stm::semanticCoherence(fit, ho$documents)),
             exclusivity = mean(stm::exclusivity(fit)))
    }, .options = furrr::furrr_options(seed = tx$seed)) |> purrr::list_rbind()
  })
  future::plan(future::sequential)
  metrics |>
    tidyr::pivot_longer(-K) |>
    ggplot(aes(K, value)) + geom_line(color = viridis(1, begin = 0.35)) +
    geom_point() + facet_wrap(~ name, scales = "free_y") + labs(y = NULL)
} else {
  if (is.null(tx$K))
    stop("tx$K is NULL and tx$run_searchK is FALSE. Re-run the config chunk, ",
         "or set tx$K before continuing.")
  glue::glue("searchK skipped (tx$run_searchK = FALSE); K fixed at {tx$K}")
}
```

# Structural topic model

Year enters the prevalence model as a spline (df `r tx$spline_df`), so topic attention is allowed to drift over the period instead of being averaged away; spectral initialization is deterministic given the data, removing initialization luck from a fit everything downstream depends on.

```{r stm-fit}
stm_fit <- cache_rds(p_out("stm_fit.rds"),
  stm::stm(stm_in$documents, stm_in$vocab, K = tx$K, init.type = "Spectral",
           prevalence = prev_form, data = meta, verbose = FALSE))

theta <- stm_fit$theta
colnames(theta) <- str_c("topic_", seq_len(tx$K))
paragraph_theta <- meta |> dplyr::bind_cols(tibble::as_tibble(theta))
```

# Modal topic assignment and its diagnostics

Each paragraph is assigned its **modal** topic, so topics partition the corpus (mutually exclusive, collectively exhaustive) and can serve as sampling strata in Script 2. This is the same move as modal class assignment in latent class analysis, and it inherits the same caveat: a modal assignment is an estimate, not an observation, and treating it as observed pushes classification error downstream. Two LCA diagnostics report how much error to expect. **Entropy $R^2$** summarizes overall assignment sharpness on a 0-1 scale; **AvePP**, the average posterior probability among a topic's own assigned paragraphs, does it per topic, with about 0.7 the conventional floor. Unlike LCA, STM supplies no classification error matrix over modal assignment, so no BCH-style correction is available here; the honest substitutes are the `min_theta` scope floor in Script 2 and its sensitivity analysis.

```{r modal}
#| results: asis
paragraph_theta <- paragraph_theta |>
  dplyr::mutate(topic     = max.col(theta, ties.method = "first"),
                theta_max = theta[cbind(dplyr::row_number(), topic)])

ent <- -sum(theta * log(theta + 1e-12))
entropy_r2 <- 1 - ent / (nrow(theta) * log(tx$K))

topic_summary <- tibble(
    topic = seq_len(tx$K),
    prevalence = colMeans(theta),
    frex = apply(stm::labelTopics(stm_fit, n = 8)$frex, 1, str_c,
                 collapse = ", "),
    prob = apply(stm::labelTopics(stm_fit, n = 8)$prob, 1, str_c,
                 collapse = ", ")) |>
  dplyr::left_join(
    paragraph_theta |>
      dplyr::group_by(topic) |>
      dplyr::summarise(n_paras = dplyr::n(), avepp = mean(theta_max)),
    by = "topic")

readr::write_csv(topic_summary, p_out("topic_summary.csv"))
readr::write_csv(paragraph_theta, p_out("paragraph_theta.csv"))

glue::glue("Entropy R2 = {round(entropy_r2, 3)}; topics with AvePP < 0.7: ",
           "{sum(topic_summary$avepp < 0.7)} of {tx$K}")
topic_summary |>
  dplyr::arrange(dplyr::desc(prevalence)) |>
  dplyr::select(topic, prevalence, n_paras, avepp) |>
  tbl_paged(caption = "Topics by prevalence, with assigned paragraph counts and AvePP (below 0.7 means fuzzy modal assignment).")
topic_summary |>
  dplyr::arrange(dplyr::desc(prevalence)) |>
  dplyr::mutate(ttl = str_c("Topic ", topic)) |>
  tbl_records("ttl", c("frex", "prob"),
              caption = "Topic terms: the evidence for checking each label.")
```

# Topic codebook

The codebook is written **once**, after the final STM, and is the naming record from here on. If `topic_codebook.csv` already exists on disk, this section reads it and makes no API calls: the file is assumed to carry your edits, and re-drafting would overwrite them. Delete the file to redraft from scratch.

Two passes, deliberately shaped:

**Pass 1, one call per topic** with the smaller model (`label` step). Each call sees only that topic's own evidence: its FREX terms, its highest-probability terms, and its three highest-theta paragraphs. It is blind to every other topic, to Year, and to stance. That blindness is the point: a model shown all topics at once starts fitting a narrative across them, and labels stop being traceable to the evidence a human can check them against.

**Pass 2, one manager call** with the larger model (`manager` step). Blind drafting has one predictable failure: two topics can come back with near-duplicate labels even though their term distributions differ. The manager sees all K labels with their FREX terms and resolves collisions. Its mandate is deliberately narrow, stated in the prompt and enforced by the schema: it may **rename only**. It cannot merge topics, split them, change a description, or reorder anything, because the topic structure is the statistical model's output and an LLM should not be permitted to override it.

The `proposition` column is written empty here. Script 2 fills it for the one topic it processes, since propositions are stance targets and you only need them for the topics you actually analyze.

```{r codebook}
#| results: asis
cb_path <- p_out("topic_codebook.csv")

if (file.exists(cb_path)) {
  codebook <- readr::read_csv(cb_path, show_col_types = FALSE)
  writeLines(glue::glue("Existing codebook found ({nrow(codebook)} topics); \\
                         no LLM calls made. Delete {basename(cb_path)} to \\
                         redraft."))
} else {
  ev <- topic_summary |>
    dplyr::left_join(
      paragraph_theta |>
        dplyr::group_by(topic) |>
        dplyr::slice_max(theta_max, n = 3) |>
        dplyr::summarise(ex = str_c(str_trunc(text, 300), collapse = "\n- "),
                         .groups = "drop"),
      by = "topic")

  # ---- Pass 1: one blind call per topic ------------------------------------
  label_sys <- build_prompt(
    role = str_c("You are a survey methodologist who names statistical topics ",
                 "so colleagues can use them."),
    task = str_c("Read this topic's distinctive terms and example paragraphs. ",
                 "Write a short label and one factual sentence describing what ",
                 "these paragraphs are about. Before answering, check that ",
                 "every word of your description is supported by the evidence ",
                 "shown."),
    rules = c("Use only the terms and paragraphs shown.",
              "Do not infer stance, sentiment, intent, or who is speaking.",
              "Do not invent statistics, dates, places, or named entities.",
              "If the evidence is mixed, describe the mixture plainly."),
    output = "A label of at most 4 words and a one-sentence description.")
  label_schema <- ellmer::type_object(
    label       = ellmer::type_string("Topic name, at most 4 words."),
    description = ellmer::type_string("One factual sentence."))
  label_chat <- make_chat("label", label_sys, tx$steps, tx$use_openrouter)

  drafts <- purrr::map(seq_len(nrow(ev)), function(i) {
    r <- call_one(label_chat,
                  str_c("Distinctive (FREX) terms: ", ev$frex[i],
                        "\nHighest-probability terms: ", ev$prob[i],
                        "\nExample paragraphs:\n- ", ev$ex[i]), label_schema)
    tibble(topic = ev$topic[i],
           label = r$label %||% NA_character_,
           description = r$description %||% NA_character_)
  }) |> purrr::list_rbind()
  if (anyNA(drafts$label))
    warning(sum(is.na(drafts$label)), " topic label calls failed; those rows ",
            "are blank in the codebook and need a name by hand.")

  # ---- Pass 2: manager resolves duplicate labels, renaming ONLY -------------
  mgr_sys <- build_prompt(
    role = str_c("You are a senior methodologist reviewing a topic codebook ",
                 "for clarity before it goes to coders."),
    task = str_c("You are shown every topic's draft label with its ",
                 "distinctive terms. Some labels are near-duplicates of each ",
                 "other even though the topics differ. Return a label for ",
                 "EVERY topic: unchanged where it is already distinct, and ",
                 "revised only where two or more labels would be confused. ",
                 "Before answering, check that no two returned labels are ",
                 "the same and that every topic number appears exactly once."),
    rules = c("Rename only. Never merge, split, drop, or reorder topics.",
              "Return the same number of topics you were given.",
              "Base any rename on the distinctive terms shown, nothing else.",
              "Keep labels to at most 4 words.",
              "Do not write descriptions; they are not yours to change."),
    output = "One label per topic number.")
  mgr_schema <- ellmer::type_object(
    topics = ellmer::type_array(
      items = ellmer::type_object(
        topic = ellmer::type_integer("The topic number, unchanged."),
        label = ellmer::type_string("Final label, at most 4 words."))))
  mgr_prompt <- str_c(
    "Draft labels with their distinctive terms:\n",
    str_c("- topic ", drafts$topic, ": \"", drafts$label, "\" | terms: ",
          ev$frex[match(drafts$topic, ev$topic)], collapse = "\n"))
  mgr <- call_one(make_chat("manager", mgr_sys, tx$steps, tx$use_openrouter),
                  mgr_prompt, mgr_schema)

  # The manager is advisory: its output is applied ONLY if it returns exactly
  # one label per topic with no topics invented or lost. Any deviation and the
  # blind drafts stand, because a malformed rename must never reshape the
  # codebook silently.
  final <- if (!is.null(mgr) && length(mgr$topics) == nrow(drafts)) {
    renamed <- purrr::map(mgr$topics, ~ tibble(topic = as.integer(.x$topic),
                                               label_new = .x$label)) |>
      purrr::list_rbind()
    if (setequal(renamed$topic, drafts$topic)) {
      drafts |> dplyr::left_join(renamed, by = "topic") |>
        dplyr::mutate(label = dplyr::coalesce(label_new, label)) |>
        dplyr::select(-label_new)
    } else drafts
  } else drafts
  n_renamed <- sum(final$label != drafts$label, na.rm = TRUE)

  codebook <- final |>
    dplyr::mutate(proposition = NA_character_) |>   # Script 2 fills this
    dplyr::left_join(topic_summary |> dplyr::select(topic, prevalence, avepp),
                     by = "topic") |>
    dplyr::arrange(topic)
  readr::write_csv(codebook, cb_path)
  writeLines(glue::glue("Codebook drafted: {nrow(codebook)} topics, \\
                         {n_renamed} renamed by the manager. EDIT \\
                         {basename(cb_path)} before running Script 2; it will \\
                         not be redrafted while the file exists."))
}

codebook |>
  dplyr::mutate(ttl = str_c("Topic ", topic, ": ", label)) |>
  tbl_records("ttl", "description",
              caption = "Topic codebook. Labels drafted blind per topic, then deduplicated by a rename-only manager pass. Edit the CSV; nothing downstream redrafts it.")
```

# Representative paragraphs

```{r thoughts}
#| results: asis
top4 <- topic_summary |> dplyr::slice_max(prevalence, n = 4) |>
  dplyr::pull(topic)
th <- stm::findThoughts(stm_fit, texts = meta$para_uid, n = 3, topics = top4)
tibble(topic = rep(top4, each = 3), para_uid = unlist(th$docs)) |>
  dplyr::left_join(paragraphs |> dplyr::select(para_uid, text),
                   by = "para_uid") |>
  dplyr::mutate(ttl = str_c("Topic ", topic, " | ", para_uid)) |>
  tbl_records("ttl", "text",
              caption = "Three highest-theta paragraphs for the four most prevalent topics.")
```

# Prevalence over time

```{r effects}
stm_year_effects <- cache_rds(p_out("stm_year_effects.rds"),
  stm::estimateEffect(
    as.formula(str_c("1:", tx$K, " ~ ", prev_rhs)),
    stm_fit, metadata = meta, uncertainty = "Global"))
```

Expected topic prevalence is predicted across the observed year range from the
fitted effect object, with 95 percent intervals. Two details make the
prediction correct rather than merely plausible. The spline basis is rebuilt at
the prediction grid with `predict(bs, ...)`, which reuses the **original
knots**; calling `ns()` afresh on the grid would place knots differently and
mis-predict silently. And the intervals come from the simulation draws
`estimateEffect` already stores under `uncertainty = "Global"`, so they carry
the topic-model uncertainty rather than regression standard errors alone.

```{r trend-estimates}
# Prediction grid: every observed year, with the basis evaluated at the
# ORIGINAL knots so the fitted coefficients apply.
grid <- tibble(Year = sort(unique(meta$Year)))
Xg   <- cbind(1, predict(bs, newx = grid$Year))

# estimateEffect stores per-topic simulation draws of the coefficient vector
# in $parameters[[k]][[j]]$est. Predicting each draw and taking quantiles
# propagates both the regression and the topic-model uncertainty.
draws_for_topic <- function(k) {
  purrr::map(stm_year_effects$parameters[[k]], function(sim)
    as.numeric(Xg %*% sim$est)) |>
    purrr::reduce(rbind) |> matrix(ncol = nrow(grid))
}
topic_trends <- purrr::map(seq_along(stm_year_effects$topics), function(i) {
  d <- draws_for_topic(i)
  grid |> dplyr::mutate(
    topic = stm_year_effects$topics[i],
    fit   = apply(d, 2, mean),
    lo    = apply(d, 2, quantile, 0.025),
    hi    = apply(d, 2, quantile, 0.975))
}) |> purrr::list_rbind()
readr::write_csv(topic_trends, p_out("topic_trends.csv"))
glue::glue("Predicted prevalence for {dplyr::n_distinct(topic_trends$topic)} \\
            topics across {nrow(grid)} years written to topic_trends.csv")
```

```{r trend-plots}
#| fig-height: 8
# All topics, paged so panels stay legible: 25 facets on one page renders each
# about 1.5 inches wide and the ribbons become unreadable.
trend_panel <- function(d, page, n_pages) {
  ggplot(d, aes(Year, fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2,
                fill = viridis(1, begin = 0.35)) +
    geom_line(linewidth = 0.7, color = viridis(1, begin = 0.35)) +
    facet_wrap(~ topic_label, scales = "free_y", ncol = 3) +
    labs(x = NULL, y = "expected topic proportion",
         title = glue::glue("Topic prevalence over time ({page} of {n_pages})"),
         subtitle = "Model-based estimates with 95 percent intervals")
}
trend_pages <- topic_trends |>
  dplyr::left_join(topic_summary |> dplyr::select(topic, prevalence),
                   by = "topic") |>
  dplyr::left_join(codebook |> dplyr::select(topic, label), by = "topic") |>
  dplyr::mutate(
    topic_label = str_c("T", topic, ": ", str_trunc(label, 22)),
    page = (dplyr::dense_rank(dplyr::desc(prevalence)) - 1L) %/%
             tx$facets_per_page + 1L) |>
  dplyr::group_split(page)
purrr::iwalk(trend_pages, function(d, i)
  print(trend_panel(d, i, length(trend_pages))))
```

```{r handoff}
glue::glue("outputs/ written: paragraphs.csv, paragraph_theta.csv (ids, Year, \\
n_tokens, text, topic_1..K, topic, theta_max), topic_summary.csv, \\
topic_codebook.csv (labels; Script 2 adds propositions), topic_trends.csv, \
stm_fit.rds, stm_year_effects.rds. Script 2 reads these.")
```
