---
title: "UNGA speeches: topics and how they change over time"
subtitle: "Step 1 of 2. Then stance.R"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

```{r config}
tx <- list(
  input_file = "unga.parquet",  # in unga/inputs/: id, para_id, year, country, raw_text
  seed = 2026L,
  min_docfreq = 5L,  # drop words in fewer paragraphs than this
  drop_words = c("united", "nations", "general", "assembly", "session",
                 "president", "mr", "secretary", "delegation"),  # in every speech
  run_searchK = FALSE,  # TRUE once to choose K (slow)
  K_grid = seq(10, 40, 5),
  K = 20L,  # set from the searchK plots
  spline_df = 5L,  # flexibility of the year trend
  n_words = 20L,  # words per topic shown to the codebook model
  n_examples = 5L,  # example paragraphs per topic shown to the codebook model
  n_draws = 25L,  # draws of topic shares behind each trend interval; memory is draws x paragraphs x K
  workers = 8L,
  use_openrouter = TRUE,  # FALSE at work
  steps = tibble::tribble(
    ~step, ~base_url_env, ~api_key_env, ~model,
    "codebook", "COMPASS_LARGE_URL", "COMPASS_LARGE_KEY", "LARGE-MODEL"))

suppressPackageStartupMessages({
  library(here); library(stm); library(splines); library(viridis)
  library(tidyverse)  # tidyverse last so its verbs win
})
theme_set(theme_minimal(base_size = 12))
set.seed(tx$seed)
# The topics project's functions, pointed at unga/inputs and unga/outputs.
options(hot_topic.project = "unga")
source(here::here("topics", "topics_source.R"))
if (isTRUE(tx$use_openrouter)) {
  source(here::here("topics", "llm_openrouter.R"))
  tx$steps$model <- open_router_models[tx$steps$step]
}
dir.create(p_out(), showWarnings = FALSE)
```

# Input

Each row is one paragraph of one speech. `id` identifies the speech, `para_id` the paragraph within it.

```{r input}
#| results: asis
raw_in <- p_in(tx$input_file)
raw <- arrow::read_parquet(raw_in)
miss <- setdiff(c("id", "para_id", "year", "country", "raw_text"), names(raw))
if (length(miss)) stop(tx$input_file, " is missing: ", str_c(miss, collapse = ", "))
paragraphs <- raw |>
  dplyr::transmute(para_uid = str_c(id, "-", para_id), id = as.character(id),
                   para_id, year = as.integer(year), country,
                   text = str_squish(raw_text)) |>
  dplyr::filter(!is.na(year), !is.na(country), !is.na(text), text != "")
stopifnot("id + para_id must be unique" = !anyDuplicated(paragraphs$para_uid),
          "a speech id must have one country and one year" =
            nrow(dplyr::distinct(paragraphs, id, country, year)) ==
            dplyr::n_distinct(paragraphs$id))
writeLines(glue::glue("Kept {nrow(paragraphs)} of {nrow(raw)} paragraphs from \\
  {dplyr::n_distinct(paragraphs$id)} speeches by \\
  {dplyr::n_distinct(paragraphs$country)} countries, \\
  {min(paragraphs$year)} to {max(paragraphs$year)}."))
```

```{r coverage}
#| fig-height: 3.5
#| fig-cap: "Countries with a speech in each year. Not every country speaks every year; country is kept in the outputs but is not in the topic model."
paragraphs |>
  dplyr::distinct(year, country) |>
  dplyr::count(year) |>
  ggplot(aes(year, n)) +
  geom_col(fill = viridis(1, begin = 0.35)) +
  labs(x = NULL, y = "countries")
```

# Document-term matrix

Words are lowercased and kept only if purely alphabetic. Stopwords, month and weekday names, and the boilerplate in `drop_words` are removed. Words in fewer than `r tx$min_docfreq` paragraphs are dropped.

```{r dtm}
tokens <- clean_tokens(paragraphs, tx$drop_words)
stm_in <- stm_input(tokens, tx$min_docfreq)
# stm needs a plain data frame in the same row order as the documents, with
# the year spline as ordinary columns yr1, yr2, ...
meta <- tibble(para_uid = names(stm_in$documents)) |>
  dplyr::left_join(paragraphs, by = "para_uid") |>
  as.data.frame()
bs <- splines::ns(meta$year, df = tx$spline_df)
colnames(bs) <- str_c("yr", seq_len(ncol(bs)))
meta <- cbind(meta, as.data.frame(bs))
prev_form <- as.formula(str_c("~ ", str_c(colnames(bs), collapse = " + ")))
glue::glue("{length(stm_in$documents)} paragraphs, {length(stm_in$vocab)} words; \\
  {nrow(paragraphs) - length(stm_in$documents)} paragraphs had no word left.")
```

# Choosing K

Take the smallest K where held-out likelihood has flattened and residual dispersion has stopped falling, then break ties on coherence and exclusivity. If the best values sit at the edge of the grid, widen it.

```{r searchK}
if (isTRUE(tx$run_searchK)) {
  future::plan(future::multisession, workers = tx$workers)
  ho <- stm::make.heldout(stm_in$documents, stm_in$vocab, seed = tx$seed)
  metrics <- cached(p_out("searchK.parquet"), raw_in,
    furrr::future_map(tx$K_grid, function(k) {
      fit <- stm::stm(ho$documents, ho$vocab, K = k, init.type = "Spectral",
                      prevalence = prev_form, data = meta, verbose = FALSE)
      tibble(K = k,
             heldout = stm::eval.heldout(fit, ho$missing)$expected.heldout,
             dispersion = stm::checkResiduals(fit, ho$documents)$dispersion,
             coherence = mean(stm::semanticCoherence(fit, ho$documents)),
             exclusivity = mean(stm::exclusivity(fit)))
    }, .options = furrr::furrr_options(seed = tx$seed)) |> purrr::list_rbind())
  future::plan(future::sequential)
  metrics |>
    tidyr::pivot_longer(-K) |>
    ggplot(aes(K, value)) +
    geom_line(color = viridis(1, begin = 0.35)) + geom_point() +
    facet_wrap(~ name, scales = "free_y") + labs(y = NULL)
} else glue::glue("searchK skipped; K = {tx$K}")
```

# Structural topic model

Year enters topic prevalence as a spline, so each topic's share can rise and fall over the period. Country is not in the model; it rides along in the outputs for filtering. Spectral initialization makes the fit reproducible.

```{r stm-fit}
stm_fit <- cached(p_out("stm_fit.rds"), raw_in,
  stm::stm(stm_in$documents, stm_in$vocab, K = tx$K, init.type = "Spectral",
           prevalence = prev_form, data = meta, verbose = FALSE))
K <- ncol(stm_fit$theta)
if (K != tx$K) stop("stm_fit.rds has K = ", K, " but tx$K = ", tx$K,
                    ". Delete stm_fit.rds, then re-render.")
theta <- stm_fit$theta
colnames(theta) <- str_c("topic_", seq_len(K))
```

# Assigning paragraphs to topics

Each paragraph is assigned its most likely topic. **Entropy R²** (0 to 1) says how sharply paragraphs sort into topics overall; **AvePP** is the average topic share among a topic's own paragraphs, with about 0.7 the usual floor.

```{r modal}
#| results: asis
paragraph_theta <- meta |>
  dplyr::select(para_uid, id, para_id, year, country, text) |>
  dplyr::bind_cols(tibble::as_tibble(theta)) |>
  dplyr::mutate(topic = max.col(theta, ties.method = "first"),
                theta_max = theta[cbind(dplyr::row_number(), topic)])
entropy_r2 <- 1 - (-sum(theta * log(theta + 1e-12))) / (nrow(theta) * log(K))

words <- stm::labelTopics(stm_fit, n = tx$n_words)
topic_summary <- tibble(
  topic = seq_len(K),
  prevalence = colMeans(theta),
  frex = apply(words$frex, 1, str_c, collapse = ", "),
  prob = apply(words$prob, 1, str_c, collapse = ", ")) |>
  dplyr::left_join(paragraph_theta |>
                     dplyr::group_by(topic) |>
                     dplyr::summarise(n_paras = dplyr::n(), avepp = mean(theta_max)),
                   by = "topic")

theta_out <- p_out("paragraph_theta.parquet")
arrow::write_parquet(paragraph_theta, theta_out)
arrow::write_parquet(topic_summary, p_out("topic_summary.parquet"))

writeLines(glue::glue("Entropy R2 = {round(entropy_r2, 3)}. \\
  Topics with AvePP under 0.7: {sum(topic_summary$avepp < 0.7, na.rm = TRUE)} of {K}."))
topic_summary |>
  dplyr::arrange(dplyr::desc(prevalence)) |>
  dplyr::select(topic, prevalence, n_paras, avepp) |>
  tbl_paged(caption = "Topics by prevalence.")
```

# Topic codebook

One call per topic, all in parallel. Each call sees only that topic's evidence: its `r tx$n_words` most distinctive (FREX) and most probable words and its `r tx$n_examples` most typical paragraphs. It drafts a label, a one-sentence description, and a candidate proposition for stance.R. The file is written once: if `topic_codebook.csv` exists it is read, not redrafted, so your edits survive. Delete it to redraft.

```{r codebook}
#| results: asis
cb_path <- p_out("topic_codebook.csv")
check_inputs(cb_path, theta_out)
if (file.exists(cb_path)) {
  codebook <- readr::read_csv(cb_path, show_col_types = FALSE)
  writeLines(glue::glue("Existing codebook read ({nrow(codebook)} topics); no calls made."))
} else {
  examples <- paragraph_theta |>
    dplyr::group_by(topic) |>
    dplyr::slice_max(theta_max, n = tx$n_examples, with_ties = FALSE) |>
    dplyr::summarise(ex = str_c("- ", str_trunc(text, 400), collapse = "\n"))
  ev <- dplyr::left_join(topic_summary, examples, by = "topic")
  sys <- build_prompt(
    role = "You are a survey methodologist who names statistical topics so colleagues can use them.",
    task = str_c("Read one topic's words and example paragraphs from speeches at the UN ",
                 "General Assembly. Write a short label, one factual sentence on what the ",
                 "paragraphs discuss, and one proposition: a single claim, in one sentence, ",
                 "that a paragraph on this topic could agree or disagree with. Check every ",
                 "word against the evidence."),
    rules = c("Use only the words and paragraphs shown.",
              "Do not infer which country is speaking or its intent.",
              "Do not invent statistics, dates, places, or names.",
              "The proposition states one side only: never 'X over Y' or 'X rather than Y'.",
              "If the evidence is mixed, say so plainly."),
    output = "A label of at most 4 words, a one-sentence description, and a one-sentence proposition.")
  type <- ellmer::type_object(
    label = ellmer::type_string("At most 4 words."),
    description = ellmer::type_string("One factual sentence."),
    proposition = ellmer::type_string("One claim, one sentence."))
  prompts <- str_c("Distinctive words: ", ev$frex, "\nMost probable words: ", ev$prob,
                   "\nExample paragraphs:\n", ev$ex)
  drafts <- ask_many(make_chat("codebook", sys, tx$steps, tx$use_openrouter),
                     prompts, type)
  codebook <- tibble(topic = ev$topic, label = drafts$label,
                     description = drafts$description,
                     proposition = drafts$proposition) |>
    dplyr::arrange(topic)
  readr::write_csv(codebook, cb_path, na = "")
  record_inputs(cb_path, theta_out)
  if (anyNA(codebook$label))
    warning(sum(is.na(codebook$label)), " topics got no label; name them by hand.")
}

codebook |>
  dplyr::left_join(dplyr::select(topic_summary, topic, frex), by = "topic") |>
  dplyr::mutate(title = str_c("Topic ", topic, ": ", label)) |>
  tbl_records("title", c("description", "proposition", "frex"),
              caption = "Topic codebook. Edit the CSV; nothing redrafts it.")
```

# Prevalence over time

For each topic, its share of each paragraph is regressed on the year spline with a fractional logit (Papke and Wooldridge 1996), so every estimated share stays between 0 and 1. The model is refit on `r tx$n_draws` draws of the topic shares and the results combined, so the 95 percent intervals include the topic model's own uncertainty and the clustering of paragraphs within a speech. Because the set of countries speaking changes from year to year, a trend mixes change within countries with change in who spoke.

```{r effects}
draws <- withr::with_seed(tx$seed, purrr::map(seq_len(tx$n_draws), function(i)
  do.call(rbind, stm::thetaPosterior(stm_fit, nsims = 1, type = "Global"))))
topic_trends <- purrr::map(seq_len(K), function(k)
  topic_trend(draws, k, bs, sort(unique(meta$year)), meta$id)) |>
  purrr::list_rbind() |>
  dplyr::rename(year = Year)
arrow::write_parquet(topic_trends, p_out("topic_trends.parquet"))
```

```{r trend-plots}
#| fig-height: 8
#| fig-cap: "Expected topic share by year with 95 percent intervals, nine topics per page, most prevalent first."
pages <- topic_trends |>
  dplyr::left_join(dplyr::select(codebook, topic, label), by = "topic") |>
  dplyr::left_join(dplyr::select(topic_summary, topic, prevalence), by = "topic") |>
  dplyr::mutate(name = str_c(topic, ". ", str_trunc(label, 24)),
                page = (dplyr::dense_rank(dplyr::desc(prevalence)) - 1) %/% 9) |>
  dplyr::group_split(page)
purrr::walk(pages, function(d) print(
  ggplot(d, aes(year, fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = viridis(1, begin = 0.35)) +
    geom_line(color = viridis(1, begin = 0.35)) +
    facet_wrap(~ name, scales = "free_y", ncol = 3) +
    labs(x = NULL, y = "expected topic share")))
```

```{r handoff}
#| results: asis
writeLines(glue::glue("Wrote paragraph_theta.parquet, topic_summary.parquet, \\
  topic_trends.parquet, topic_codebook.csv, stm_fit.rds to unga/outputs/. \\
  Next: pick a topic and a proposition from the codebook and run stance.R."))
```
