---
title: "Preprocessing: comments to paragraph units"
subtitle: "Step 1 of 2. Then topics.qmd"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

```{r config}
tx <- list(
  min_tokens = 60L,  # a unit needs at least this many words for the topic model
  max_tokens = 150L,  # a merge may not grow a unit past this; keeps topics apart
  min_sim_quantile = 0.20,  # pairs less similar than this share of all neighbors never merge
  window_n = 300L,  # embedding requests per minute (limit is 500)
  window_wait = 60L,  # seconds to rest between windows
  use_openrouter = TRUE,  # FALSE at work
  steps = tibble::tribble(
    ~step, ~base_url_env, ~api_key_env, ~model,
    "embed", "COMPASS_EMBED_URL", "COMPASS_EMBED_KEY", "MODEL-FOR-EMBEDDING"))

suppressPackageStartupMessages({
  library(here); library(viridis)
  library(tidyverse)  # tidyverse last so its verbs win
})
theme_set(theme_minimal(base_size = 12))
options(hot_topic.project = "topics")  # paths: topics/inputs, topics/outputs
source(here::here("topics", "topics_source.R"))
if (isTRUE(tx$use_openrouter)) {
  source(here::here("topics", "llm_openrouter.R"))
  tx$steps$model <- open_router_models[tx$steps$step]
}
dir.create(p_out(), showWarnings = FALSE)
```

# Corpus

Needs three columns: `atom_id` (comment id), `Year` (integer), `body_english` (text). The default below is a demonstration corpus that ships with quanteda; at work replace it with one line reading your file.

```{r corpus}
corpus_raw <- tibble(
  atom_id = quanteda::docnames(quanteda::data_corpus_inaugural),
  Year = quanteda::docvars(quanteda::data_corpus_inaugural, "Year"),
  body_english = as.character(quanteda::data_corpus_inaugural))
# At work:
# corpus_raw <- readr::read_csv(p_in("YOUR_FILE.csv"))
```

```{r validate}
#| results: asis
miss <- setdiff(c("atom_id", "Year", "body_english"), names(corpus_raw))
if (length(miss)) stop("corpus_raw is missing: ", str_c(miss, collapse = ", "))
stopifnot("atom_id must be unique" = !anyDuplicated(corpus_raw$atom_id),
          "Year must be whole numbers" = all(corpus_raw$Year == round(corpus_raw$Year), na.rm = TRUE))
n0 <- nrow(corpus_raw)
corpus_raw <- dplyr::filter(corpus_raw, !is.na(Year), !is.na(body_english),
                            str_squish(body_english) != "")
writeLines(glue::glue("Kept {nrow(corpus_raw)} of {n0} comments, \\
  {min(corpus_raw$Year)} to {max(corpus_raw$Year)}."))
```

# From lines to paragraph units

Most comments here break after every sentence, so splitting at line breaks gives orphan sentences too short for a topic model. Three steps rebuild usable units.

1. **Split and repair.** Text is split at line breaks. A line that follows a line without end punctuation and starts lowercase is a wrapped piece of the same sentence and is joined back.
2. **Merge by meaning.** Every comment is embedded, one vector per block. Within a comment, the most similar pair of *neighboring* blocks is merged first, as long as one of them is shorter than `min_tokens`, the result stays within `max_tokens`, and the two are at least as similar as the `min_sim_quantile` point of all neighboring pairs in the corpus. This repeats until no pair qualifies. Reading order is never broken.
3. **Drop what cannot be placed.** A unit still under `min_tokens` had no neighbor it was similar enough to join without breaking the size ceiling. It is left out of the topic model rather than forced into a paragraph it would distort, and written to `dropped_units.csv` so you can read what was lost.

The two token settings are the balance you are choosing: `min_tokens` is enough text for the topic model to read, `max_tokens` is the point past which a unit starts to hold more than one subject. `topics.qmd` refits the topic model at other settings to show whether that choice drives the topics.

```{r blocks}
#| results: asis
blocks <- split_blocks(corpus_raw)
tibble(measure = c("lines", "blocks after rejoining wrapped lines",
                   "median words per block", "blocks under min_tokens"),
       value = c(attr(blocks, "lines_in"), nrow(blocks),
                 stats::median(blocks$n_tokens),
                 sum(blocks$n_tokens < tx$min_tokens))) |>
  tbl_paged(caption = "Blocks before merging.")
```

```{r embed}
emb <- embed_all(blocks, p_out("embeddings.parquet"), tx$steps, tx$use_openrouter,
                 tx$window_n, tx$window_wait)
```

```{r link}
#| results: asis
sims <- adjacent_sims(blocks, emb$vec)
min_sim <- stats::quantile(sims, tx$min_sim_quantile, names = FALSE)
units <- link_corpus(blocks, emb$vec, tx$min_tokens, tx$max_tokens, min_sim)
paragraphs <- dplyr::filter(units, keep)
dropped <- dplyr::filter(units, !keep)

tibble(measure = c("similarity threshold (cosine)", "units formed",
                   "units kept", "units dropped", "words dropped (share)",
                   "comments with nothing kept", "median words per kept unit",
                   "median blocks per kept unit"),
       value = c(round(min_sim, 3), nrow(units), nrow(paragraphs), nrow(dropped),
                 round(sum(dropped$n_tokens) / sum(units$n_tokens), 3),
                 dplyr::n_distinct(corpus_raw$atom_id) - dplyr::n_distinct(paragraphs$atom_id),
                 stats::median(paragraphs$n_tokens), stats::median(paragraphs$n_blocks))) |>
  tbl_paged(caption = "Merging results. Watch the share of words dropped and the comments with nothing kept: if either is large, lower min_tokens or min_sim_quantile.")
```

```{r units-plot}
#| fig-cap: "Words per kept unit. Units pile up between min_tokens and max_tokens (dashed); long author paragraphs sit to the right and are never split."
ggplot(paragraphs, aes(n_tokens)) +
  geom_histogram(bins = 40, fill = viridis(1, begin = 0.35)) +
  geom_vline(xintercept = c(tx$min_tokens, tx$max_tokens), linetype = 2) +
  labs(x = "words per unit", y = "units")
```

```{r dropped-plot}
#| fig-cap: "Share of words dropped by year. Dropping is not neutral if it concentrates in some years: those years then rest on longer, better-formed comments."
units |>
  dplyr::group_by(Year) |>
  dplyr::summarise(dropped = sum(n_tokens[!keep]) / sum(n_tokens)) |>
  ggplot(aes(Year, dropped)) +
  geom_col(fill = viridis(1, begin = 0.6)) +
  labs(x = NULL, y = "share of words dropped")
```

# Handoff

```{r handoff}
#| results: asis
readr::write_csv(dplyr::select(paragraphs, para_uid, atom_id, para_id, Year,
                               n_tokens, n_blocks, text), p_out("paragraphs.csv"))
readr::write_csv(dplyr::select(dropped, atom_id, Year, n_tokens, text),
                 p_out("dropped_units.csv"))
readr::write_csv(blocks, p_out("blocks.csv"))
arrow::write_parquet(tibble(min_tokens = tx$min_tokens, max_tokens = tx$max_tokens,
                            min_sim = min_sim), p_out("merge_settings.parquet"))
writeLines(glue::glue("Wrote paragraphs.csv ({nrow(paragraphs)} units), \\
  dropped_units.csv ({nrow(dropped)}), blocks.csv, merge_settings.parquet, \\
  embeddings.parquet. Next: topics.qmd."))
```
