---
title: "Topics and how they change over time"
subtitle: "Step 2 of 2. Reads preprocess.qmd outputs"
format:
  typst:
    toc: false
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

```{r config}
tx <- list(
  seed = 2026L,
  min_docfreq = 5L,  # drop words in fewer paragraphs than this
  run_searchK = FALSE,  # TRUE once to choose K (slow)
  K_grid = seq(5, 30, 3),
  K = 12L,  # set from the searchK plots
  spline_df = 5L,  # flexibility of the year trend
  run_sensitivity = FALSE,  # TRUE once to test the merge settings
  sens_factors = c(0.75, 1.25),  # multiples of min_tokens and max_tokens
  n_words = 20L,  # words per topic shown to the codebook model
  n_examples = 5L,  # example paragraphs per topic shown to the codebook model
  n_draws = 25L,  # draws of topic shares behind each trend interval
  workers = 8L,
  use_openrouter = TRUE,  # FALSE at work
  steps = tibble::tribble(
    ~step, ~base_url_env, ~api_key_env, ~model,
    "codebook", "COMPASS_LARGE_URL", "COMPASS_LARGE_KEY", "LARGE-MODEL"))

suppressPackageStartupMessages({
  library(here); library(stm); library(splines); library(viridis)
  library(tidyverse)  # tidyverse last so its verbs win
})
theme_set(theme_minimal(base_size = 12))
set.seed(tx$seed)
source(here::here("topics", "topics_source.R"))
if (isTRUE(tx$use_openrouter)) {
  source(here::here("topics", "llm_openrouter.R"))
  tx$steps$model <- open_router_models[tx$steps$step]
}
```

# Input

```{r input}
#| results: asis
para_in <- p_out("paragraphs.csv")
paragraphs <- readr::read_csv(para_in, show_col_types = FALSE)
writeLines(glue::glue("{nrow(paragraphs)} paragraph units from \\
  {dplyr::n_distinct(paragraphs$atom_id)} comments."))
```

# Document-term matrix

Words are lowercased and kept only if purely alphabetic, which drops numbers and special characters. Stopwords and month and weekday names are removed; temporal words would let topics encode the year directly and leak time into the trends. Words in fewer than `r tx$min_docfreq` paragraphs are dropped.

```{r dtm}
tokens <- clean_tokens(paragraphs)
stm_in <- stm_input(tokens, tx$min_docfreq)
# stm needs a plain data frame (not a tibble) in the same row order as the
# documents, and the year spline as ordinary columns yr1, yr2, ...
meta <- tibble(para_uid = names(stm_in$documents)) |>
  dplyr::left_join(paragraphs, by = "para_uid") |>
  as.data.frame()
bs <- splines::ns(meta$Year, df = tx$spline_df)
colnames(bs) <- str_c("yr", seq_len(ncol(bs)))
meta <- cbind(meta, as.data.frame(bs))
prev_rhs <- str_c(colnames(bs), collapse = " + ")
prev_form <- as.formula(str_c("~ ", prev_rhs))
glue::glue("{length(stm_in$documents)} paragraphs, {length(stm_in$vocab)} words")
```

# Choosing K

Take the smallest K where held-out likelihood has flattened and residual dispersion has stopped falling, then break ties on coherence and exclusivity. If the best values sit at the edge of the grid, widen the grid.

```{r searchK}
if (isTRUE(tx$run_searchK)) {
  future::plan(future::multisession, workers = tx$workers)
  ho <- stm::make.heldout(stm_in$documents, stm_in$vocab, seed = tx$seed)
  metrics <- cached(p_out("searchK.parquet"), para_in,
    furrr::future_map(tx$K_grid, function(k) {
      fit <- stm::stm(ho$documents, ho$vocab, K = k, init.type = "Spectral",
                      prevalence = prev_form, data = meta, verbose = FALSE)
      tibble(K = k,
             heldout = stm::eval.heldout(fit, ho$missing)$expected.heldout,
             dispersion = stm::checkResiduals(fit, ho$documents)$dispersion,
             coherence = mean(stm::semanticCoherence(fit, ho$documents)),
             exclusivity = mean(stm::exclusivity(fit)))
    }, .options = furrr::furrr_options(seed = tx$seed)) |> purrr::list_rbind())
  future::plan(future::sequential)
  metrics |>
    tidyr::pivot_longer(-K) |>
    ggplot(aes(K, value)) +
    geom_line(color = viridis(1, begin = 0.35)) + geom_point() +
    facet_wrap(~ name, scales = "free_y") + labs(y = NULL)
} else glue::glue("searchK skipped; K = {tx$K}")
```

# Structural topic model

Year enters topic prevalence as a spline, so each topic's share can rise and fall over the period. Spectral initialization makes the fit reproducible. The fit is saved as `stm_fit.rds` and reloaded on later renders; delete it to refit.

```{r stm-fit}
stm_fit <- cached(p_out("stm_fit.rds"), para_in,
  stm::stm(stm_in$documents, stm_in$vocab, K = tx$K, init.type = "Spectral",
           prevalence = prev_form, data = meta, verbose = FALSE))
K <- ncol(stm_fit$theta)
if (K != tx$K) stop("stm_fit.rds has K = ", K, " but tx$K = ", tx$K,
                    ". Delete stm_fit.rds, then re-render.")
theta <- stm_fit$theta
colnames(theta) <- str_c("topic_", seq_len(K))
```

# Assigning paragraphs to topics

Each paragraph is assigned its most likely topic. **Entropy R²** (0 to 1) says how sharply paragraphs sort into topics overall; **AvePP** is the average topic share among a topic's own paragraphs, with about 0.7 the usual floor. Both fall as units get longer, because longer units genuinely mix topics, so use them to compare topics within a fit, not to choose the unit size.

```{r modal}
#| results: asis
paragraph_theta <- meta |>
  dplyr::select(para_uid, atom_id, para_id, Year, n_tokens, text) |>
  dplyr::bind_cols(tibble::as_tibble(theta)) |>
  dplyr::mutate(topic = max.col(theta, ties.method = "first"),
                theta_max = theta[cbind(dplyr::row_number(), topic)])
entropy_r2 <- 1 - (-sum(theta * log(theta + 1e-12))) / (nrow(theta) * log(K))

words <- stm::labelTopics(stm_fit, n = tx$n_words)
topic_summary <- tibble(
  topic = seq_len(K),
  prevalence = colMeans(theta),
  frex = apply(words$frex, 1, str_c, collapse = ", "),
  prob = apply(words$prob, 1, str_c, collapse = ", ")) |>
  dplyr::left_join(paragraph_theta |>
                     dplyr::group_by(topic) |>
                     dplyr::summarise(n_paras = dplyr::n(), avepp = mean(theta_max)),
                   by = "topic")

readr::write_csv(paragraph_theta, p_out("paragraph_theta.csv"))
readr::write_csv(topic_summary, p_out("topic_summary.csv"))
theta_in <- p_out("paragraph_theta.csv")

writeLines(glue::glue("Entropy R2 = {round(entropy_r2, 3)}. \\
  Topics with AvePP under 0.7: {sum(topic_summary$avepp < 0.7, na.rm = TRUE)} of {K}."))
topic_summary |>
  dplyr::arrange(dplyr::desc(prevalence)) |>
  dplyr::select(topic, prevalence, n_paras, avepp) |>
  tbl_paged(caption = "Topics by prevalence.")
```

# Do the merge settings drive the topics?

The paragraph units are rebuilt at `r str_c(tx$sens_factors, collapse = " and ")` times `min_tokens` and `max_tokens`, from the embeddings already saved (no API calls), and the model is refit at the same K. Each baseline topic is matched to its closest topic in each refit by the cosine of their word distributions. High cosines mean the topics belong to the corpus; a topic with a low cosine exists only at one way of cutting the text and deserves caution.

```{r sensitivity}
#| results: asis
if (isTRUE(tx$run_sensitivity)) {
  sens_in <- p_out(c("paragraphs.csv", "blocks.csv", "embeddings.parquet",
                     "merge_settings.parquet"))
  blocks <- readr::read_csv(p_out("blocks.csv"), show_col_types = FALSE)
  emb <- emb_read(p_out("embeddings.parquet"))
  ms <- arrow::read_parquet(p_out("merge_settings.parquet"))
  refit <- function(f) {
    p <- link_corpus(blocks, emb$vec, ms$min_tokens * f, ms$max_tokens * f,
                     ms$min_sim) |>
      dplyr::filter(keep)
    s <- stm_input(clean_tokens(p), tx$min_docfreq)
    fit <- stm::stm(s$documents, s$vocab, K = K, init.type = "Spectral",
                    verbose = FALSE)
    match_topics(beta_matrix(stm_fit), beta_matrix(fit)) |>
      dplyr::mutate(factor = f, paragraphs = length(s$documents))
  }
  sens_tbl <- cached(p_out("sensitivity.parquet"), sens_in,
                     purrr::list_rbind(purrr::map(tx$sens_factors, refit)))
  if (max(sens_tbl$topic) != K)
    stop("sensitivity.parquet was built at another K. Delete it and re-render.")
  sens_tbl |>
    dplyr::group_by(factor) |>
    dplyr::summarise(paragraphs = dplyr::first(paragraphs),
                     median_cosine = stats::median(cosine),
                     lowest_cosine = min(cosine)) |>
    tbl_paged(caption = "Matched topic cosine against the baseline fit.")
  sens_tbl |>
    tidyr::pivot_wider(id_cols = topic, names_from = factor,
                       values_from = cosine, names_prefix = "x") |>
    tbl_paged(caption = "Matched cosine per topic.")
} else writeLines("Sensitivity test skipped (tx$run_sensitivity = FALSE).")
```

# Topic codebook

One call per topic, all in parallel, with the larger model. Each call sees only that topic's evidence: its `r tx$n_words` most distinctive (FREX) and `r tx$n_words` most probable words and its `r tx$n_examples` most typical paragraphs. It drafts a label, a one-sentence description, and a candidate proposition: a claim, in the form the stance app wants, that a reader could paste in to measure where this topic stands. The file is written once: if `topic_codebook.csv` exists it is read, not redrafted, so your edits survive. Delete it to redraft.

```{r codebook}
#| results: asis
cb_path <- p_out("topic_codebook.csv")
check_inputs(cb_path, theta_in)
if (file.exists(cb_path)) {
  codebook <- readr::read_csv(cb_path, show_col_types = FALSE)
  writeLines(glue::glue("Existing codebook read ({nrow(codebook)} topics); no calls made."))
} else {
  examples <- paragraph_theta |>
    dplyr::group_by(topic) |>
    dplyr::slice_max(theta_max, n = tx$n_examples, with_ties = FALSE) |>
    dplyr::summarise(ex = str_c("- ", str_trunc(text, 400), collapse = "\n"))
  ev <- dplyr::left_join(topic_summary, examples, by = "topic")
  sys <- build_prompt(
    role = "You are a survey methodologist who names statistical topics so colleagues can use them.",
    task = str_c("Read one topic's words and example paragraphs from public comments. Write a ",
                 "short label, one factual sentence on what the paragraphs discuss, and one ",
                 "proposition: a single claim, in one sentence, that a paragraph on this ",
                 "topic could agree or disagree with. Check every word against the evidence."),
    rules = c("Use only the words and paragraphs shown.",
              "Do not infer who is speaking or their intent.",
              "Do not invent statistics, dates, places, or names.",
              "The proposition states one side only: never 'X over Y' or 'X rather than Y'.",
              "If the evidence is mixed, say so plainly."),
    output = "A label of at most 4 words, a one-sentence description, and a one-sentence proposition.")
  type <- ellmer::type_object(
    label = ellmer::type_string("At most 4 words."),
    description = ellmer::type_string("One factual sentence."),
    proposition = ellmer::type_string("One claim, one sentence."))
  prompts <- str_c("Distinctive words: ", ev$frex, "\nMost probable words: ", ev$prob,
                   "\nExample paragraphs:\n", ev$ex)
  drafts <- ask_many(make_chat("codebook", sys, tx$steps, tx$use_openrouter),
                     prompts, type)
  codebook <- tibble(topic = ev$topic, label = drafts$label,
                     description = drafts$description,
                     proposition = drafts$proposition) |>
    dplyr::arrange(topic)
  readr::write_csv(codebook, cb_path, na = "")
  record_inputs(cb_path, theta_in)
  if (anyNA(codebook$label))
    warning(sum(is.na(codebook$label)), " topics got no label; name them by hand.")
  writeLines(glue::glue("Codebook drafted for {nrow(codebook)} topics. Edit \\
    topic_codebook.csv; it is not redrafted while it exists."))
}

dup <- codebook |> dplyr::add_count(label) |> dplyr::filter(n > 1, !is.na(label))
if (nrow(dup)) writeLines(glue::glue("**Duplicate labels to fix:** \\
  {str_c('topic ', dup$topic, ' (', dup$label, ')', collapse = ', ')}."))

codebook |>
  dplyr::left_join(dplyr::select(topic_summary, topic, frex), by = "topic") |>
  dplyr::mutate(title = str_c("Topic ", topic, ": ", label)) |>
  tbl_records("title", c("description", "proposition", "frex"),
              caption = "Topic codebook. Edit the CSV; nothing redrafts it.")
```

# Prevalence over time

For each topic, its share of each paragraph is regressed on the year spline with a fractional logit (a quasibinomial model for proportions; Papke and Wooldridge 1996), so every estimated share stays between 0 and 1. This replaces `stm::estimateEffect`, whose linear model can predict negative shares for rare topics. The model is refit on `r tx$n_draws` draws of the topic shares from the fitted STM and the results are combined, so the 95 percent intervals include the topic model's own uncertainty as well as the fact that paragraphs from one comment are related.

```{r effects}
draws <- withr::with_seed(tx$seed, purrr::map(seq_len(tx$n_draws), function(i)
  do.call(rbind, stm::thetaPosterior(stm_fit, nsims = 1, type = "Global"))))
grid <- sort(unique(meta$Year))
topic_trends <- purrr::map(seq_len(K), function(k)
  topic_trend(draws, k, bs, grid, meta$atom_id)) |>
  purrr::list_rbind()
readr::write_csv(topic_trends, p_out("topic_trends.csv"))
```

```{r trend-plots}
#| fig-height: 8
#| fig-cap: "Expected topic share by year with 95 percent intervals, nine topics per page, most prevalent first."
pages <- topic_trends |>
  dplyr::left_join(dplyr::select(codebook, topic, label), by = "topic") |>
  dplyr::left_join(dplyr::select(topic_summary, topic, prevalence), by = "topic") |>
  dplyr::mutate(name = str_c(topic, ". ", str_trunc(label, 24)),
                page = (dplyr::dense_rank(dplyr::desc(prevalence)) - 1) %/% 9) |>
  dplyr::group_split(page)
purrr::walk(pages, function(d) print(
  ggplot(d, aes(Year, fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = viridis(1, begin = 0.35)) +
    geom_line(color = viridis(1, begin = 0.35)) +
    facet_wrap(~ name, scales = "free_y", ncol = 3) +
    labs(x = NULL, y = "expected topic share")))
```

```{r handoff}
#| results: asis
writeLines(glue::glue("Wrote paragraph_theta.csv, topic_summary.csv, \\
  topic_codebook.csv, topic_trends.csv, stm_fit.rds to topics/outputs/. \\
  The app reads the four CSVs. To measure stance on one of the propositions \\
  above, copy paragraph_theta.csv into stance/inputs/ and read stance.qmd."))
```
